---
title: 'Practical Machine Learning: Course Project'
author: "by_mcmillan03"
subtitle: Analysis of Weight Lifting Exercise Dataset
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). "

```{r init, echo=TRUE}
library(caret)
library(rpart.plot)
library(rattle)
library(randomForest)
```

## Loading and preprocessing the data

Set working directory to parent directory of data. Load the data locally or from the web.

```{r load_data, echo=TRUE}
setwd("C:/Users/smcmillan/Documents/Online\ Courses/JHU_08_Practical_Machine_Learning/Assignments/Project/PracticalML")

training_file = "pml-training.csv"
if (file.exists(training_file)) {
  pml_training_data = read.csv(training_file, stringsAsFactors = TRUE, na.strings=c("NA",""), header = TRUE)
} else {
  pml_training_data = read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"),header = TRUE)
  write.csv(pml_training_data, training_file)
}
```

Select the columns and rows that will be used for training and validation.  Note that many columns have mostly NA (the corrsesponding columns in the testing data are ALL NA).  And split into training and testing sets.  Note that randomForests are used so a smaller training set is chosen.

```{r subset, echo=TRUE}
# Only choose the columns with enough "information" (the testing set has zero information for the others, actually)
colnames = c("user_name",
             #"raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp",
             "new_window", #Always "no" in final_testing
             "num_window",
             "roll_belt", "pitch_belt", "yaw_belt", "total_accel_belt",
             "gyros_belt_x", "gyros_belt_y", "gyros_belt_z",
             "accel_belt_x", "accel_belt_y", "accel_belt_z",
             "magnet_belt_x", "magnet_belt_y", "magnet_belt_z",
             "roll_arm", "pitch_arm", "yaw_arm", "total_accel_arm",
             "gyros_arm_x", "gyros_arm_y", "gyros_arm_z",
             "accel_arm_x", "accel_arm_y", "accel_arm_z",
             "magnet_arm_x", "magnet_arm_y", "magnet_arm_z",
             "roll_dumbbell", "pitch_dumbbell", "yaw_dumbbell", "total_accel_dumbbell",
             "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z",
             "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z",
             "magnet_dumbbell_x", "magnet_dumbbell_y", "magnet_dumbbell_z",
             "roll_forearm", "pitch_forearm", "yaw_forearm", "total_accel_forearm",
             "gyros_forearm_x", "gyros_forearm_y", "gyros_forearm_z",
             "accel_forearm_x", "accel_forearm_y", "accel_forearm_z",
             "magnet_forearm_x", "magnet_forearm_y", "magnet_forearm_z",
             "classe"
             )

# Note: new_window field in final testing data is "no" for all records.  Chose same records here.
tmp_data = subset(pml_training_data[,colnames], new_window=="no")
#summary(tmp_data)

# split into training and testing sets.
set.seed(8088)
inTest = createDataPartition(y=tmp_data$classe, p = 0.7, list=FALSE)
training = tmp_data[-inTest,]
testing  = tmp_data[ inTest,]
#str(training)
#summary(training)
#names(training)
#head(training)
```

## Decision Tree Classification

```{r decision_tree_model, echo=TRUE}
tree_model_file = "my_tree_model.rds"
if (file.exists(tree_model_file)) {
  tree_model = readRDS(tree_model_file)
} else {
  tree_model = train(classe~., data=training, method="rpart")
  saveRDS(tree_model, "my_tree_model.rds")
}

summary(tree_model)

fancyRpartPlot(tree_model$finalModel)
predictions = predict(tree_model, newdata=testing)
confusionMatrix(predictions, testing$classe)
```

### Random Forest Classification

```{r random_forest_model, echo=TRUE}
rf_model_file = "my_rf_model.rds"
if (file.exists(rf_model_file)){
  rf_model = readRDS(rf_model_file)
} else {
  rf_model = train(classe~., data=testing, method="rf",prox=TRUE)
  saveRDS(rf_model, rf_model_file)
}
rf_model
getTree(rf_model$finalModel, k=2)

```

## Predicting Test Cases

```{r final_tests, echo=TRUE}
#final_testing = read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"), header = TRUE)
final_testing = read.csv("../pml-testing.csv", stringsAsFactors = TRUE, na.strings=c("NA",""), header = TRUE)
#final_testing = subset(pml_testing[,colnames], new_window=="no")
#summary(final_testing)
final_predictions = predict(rf_model, newdata=final_testing)
final_predictions
```

## Outputting Test Case Predictions

```{r write_output_files, echo=TRUE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("../test_predictions/problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

# Correct answers: B A B A A E D B A A B C B A E E A B B B
pml_write_files(final_predictions)
```

## Document Environment

```{r env, echo=TRUE}
sessionInfo()
```